---
title: 'Rejection sampling: Non-conjugate Binomial example'
author: "Patrick Graham"
date: "7 September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
library(MASS)          
library(MCMCpack)
library(arm)   #gives access to logit and inverse logit functions
library(logitnorm)  
library(LearnBayes) ##access to laplace approximation etc
library(logitnorm)
#gives access to logit - normal density function

#wd <- "~/Patrick/Stat314-2018/code" ##change to own folder or better
                                    ##still set this up as project in 
                                    ##Rsstudio

#setwd(wd)
```
## Illustration of the likelihood as a function of the model parameter.
Before looking at  rejection sampling we illustrate the idea of the 
likelihood as function of the model parameter.
Consider a study comprising 6 Binomial experiments each with N=10 trial
The data is:

```{r,echo=TRUE}
Y <- c(6,7,5,5,4,8)
N <- rep(10,length(Y) )
```
We compute some summaries typically involved in Bayesian calculations.
```{r,echo=TRUE}
NmY <- N-Y

sumY <- sum(Y)
sumNmY <- sum(N-Y)
sumY
sumNmY
```
Now we set-up a function to compute the log-likelihood. When doing numerical work with likelihoods it usually best to work on the (natural)
log scale.
```{r, echo=TRUE}

loglike_binom <- function(theta,Y,N)  {
  logl <- dbinom(Y,N,prob=theta,log=TRUE)
  return(sum(logl))
}

##Evaluate at a few points
loglike_binom(theta=0.4,Y=Y,N=N)
loglike_binom(theta=0.1,Y=Y,N=N)
loglike_binom(theta=0.8,Y=Y,N=N)
###Plot the log-likelihood for a range of sucesss probabilities

testtheta <- seq(from=0.01,to=0.99,by=0.01) #range of theta's to test

##set up structure to store log-likelihoods for plotting
loglstore <- vector(mode="numeric",length=length(testtheta))

##Loop over theta values and compute the corresponding log-likelihood

for (i in 1:length(testtheta)) {
  loglstore[i] <- loglike_binom(theta=testtheta[i],Y=Y,N=N)
}
loglstore

###Plot the log-likelihood

plot(testtheta,loglstore,main="log-likelihood for Binomial example")

##clearer if we plot a line? A continuous function after all

plot(testtheta,loglstore,type="l",lwd=2.0,main="log-likelihood for Binomial example")
```

##Rejection sampler for a non-conjugte Binomial problem

$$
\begin{aligned}
Y_{i} &  \stackrel{\mathrm{indep}}{\sim} \mathrm{Binomial}(10,\theta) \nonumber \\
\mathrm{logit}(\theta) & \sim \mathrm{Normal}(\mu,\sigma^{2}) \\
p(\theta|\mathbf{Y},N) & \propto \prod_{i}\mathrm{Binomial}(Y_{i}|10,\theta) \times \mathrm{logitN}(\theta|\mu,\sigma^{2})
\end{aligned}
$$
To get familiar with the logit-normal prior we plot this density for a few different parameter combinations.
```{r, echo=TRUE}

##first define a function to simulate from the logit-normal
rlogitnormal <- function(n,mu,sigma) {
  require(arm)
  l <- rnorm(n=n,mean=mu,sd=sigma)
  theta <- invlogit(l)
  return(theta)
}

##specify parameters of the prior
mu_prior <- logit(0.5)
sigma_prior <- 5

###Plot the prior
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta,from=0,to=1)
plot(density_prior,main=paste("logitN(",round(mu_prior,2),",",sigma_prior,")",sep=""))
summary(testtheta)

###Try another set of parameters
mu_prior <- logit(0.5)
sigma_prior <- 1

###Plot the prior
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta,from=0,to=1)
plot(density_prior,main=paste("logitN(",round(mu_prior,2),",",sigma_prior,")",sep=""))
summary(testtheta)
```

###Rejection sampling.

If our approximation is the prior, $r(\theta)$ is the likelihood. 
Hence we accept draws with probability Likelihood/M
where M is the value of the likelihood evaluated at the MLE. It is easier to work on the log-scale and so find $\log(M)$ by computing the log-likelihood with $\theta$ set to the MLE, which is $\hat{\theta}=\sum_{i}Y_{i} /\sum_{i}N_{i}.$
```{r,echo=TRUE}
logM <- loglike_binom(theta=(sum(Y)/sum(N)),Y=Y,N=N) 

##specify the prior
mu_prior <- logit(0.5)
sigma_prior <- 5

##set-up the rejection sampling loop
simnum <- 2000 #desired simulation sample size
storetheta <- vector(mode="numeric",length=simnum)
accepted <- 0
rejected <- 0

while(accepted < simnum) {
  theta_prop <- rlogitnormal(n=1,mu=mu_prior,sigma=sigma_prior)
  ###Evaluate the likelihood at this proposal
  llike_prop <- loglike_binom(Y=Y,N=N,theta=theta_prop)
  
  
  U <- runif(1)
  if (log(U) < llike_prop - logM) {
    accepted <- accepted + 1
    storetheta[accepted] <- theta_prop
  } else {
    rejected <- rejected + 1
  }
}
summary(storetheta)
plot(density(storetheta))
accepted
rejected

acceptance_rate <- accepted/(accepted+rejected)
acceptance_rate

###plot posterior density plot with prior superimposed 
#- to illustrate ideas around quality of 
##approximatin density
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta,from=0,to=1)
plot(density(storetheta),
     main="rejection sampler logit-normal(0,5) prior",lwd=2,xlim=c(0,1)) 
lines(density_prior,lwd=2,col="red")
```

Now we repeat  all this with an alternative alternative prior, which is less diffuse  and again use the prior as the approximating density.

```{r, echo=TRUE}
##specify parameters of the prior
mu_prior <- logit(0.5)  #logit(0.5) =0
sigma_prior <- 1

###Plot the prior
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta,from=0,to=1)
plot(density_prior,main=paste("logitN(",round(mu_prior,2),",",sigma_prior,")",sep=""))
summary(testtheta)
##Find log(M) - log of maximum value of the ratio of the unnormalized posterior density (q) to the approximating density
logM <- loglike_binom(theta=(sum(Y)/sum(N)),Y=Y,N=N) ##working with the
#unnormalized density
simnum <- 2000 #desired simulation sample size
storetheta <- vector(mode="numeric",length=simnum)
accepted <- 0
rejected <- 0

while(accepted < simnum) {
  theta_prop <- rlogitnormal(n=1,mu=mu_prior,sigma=sigma_prior)
  ###Evaluate the likelihood at this proposal
  llike_prop <- loglike_binom(Y=Y,N=N,theta=theta_prop)
  
  
  U <- runif(1)
  if (log(U) < llike_prop - logM) {
    accepted <- accepted + 1
    storetheta[accepted] <- theta_prop
  } else {
    rejected <- rejected + 1
  }
}
summary(storetheta)
plot(density(storetheta))
accepted
rejected

acceptance_rate <- accepted/(accepted+rejected)
acceptance_rate

###plot posterior density plot with prior superimposed 
#- to illustrate ideas around quality of 
##approximatin density
plot(density(storetheta),
     main="rejection sampler logit-normal(0,1) prior",lwd=2,xlim=c(0,1)) 
lines(density_prior,lwd=2,col="red")
```

###Using a Beta approximation to the posterior as the approximating density.

We consider again the diffuse logit-normal prior.

```{r,echo=TRUE}
##Go back to the original prior
mu_prior <- logit(0.5)
sigma_prior <- 5
##Re-draw the logit-normal  sample from the prior
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta)
plot(density(testtheta,from=0,to=1),main=paste("logitN(",round(mu_prior,2),",",sigma_prior,")",sep=""))
summary(testtheta)
```

We can approximate the logit-normal prior with a Beta distribution with the same mean and variance.

```{r, echo=TRUE}
mean_theta <- mean(testtheta)
var_theta <- (sd(testtheta))^2
mean_theta
var_theta

a <- (mean_theta^2*(1-mean_theta) - var_theta*mean_theta)/var_theta 
b <- a*(1-mean_theta)/mean_theta
a
b

###plot the beta density with these values 
##could  do  exactly using dbeta but will simulate

apprxtheta <- rbeta(10000,a,b)
summary(apprxtheta)
mean(apprxtheta)
(sd(apprxtheta))^2
density_prior_beta <- density(apprxtheta,from=0,to=1)

plot(density_prior_beta,main=paste("beta(",round(a,2),",",round(b,2),")",sep=""))
```

We can construct a beta approximation to the posterior  using standard the standard conjugate udating formulae for the binomial-beta model.

```{r,echo=TRUE}

apos <- a + sumY #sum_i Y_i
bpos <- b + sumNmY  #sum_i (N_i-Y_i)

#check approximate posterior mean
apos/(apos+bpos)

pos_approx <- rbeta(10000,apos,bpos)
density_pos_betaapprx <- density(pos_approx)
plot(density(pos_approx))
```

We Need to find the maximum  of 
$\mathrm{Binomial}(Y|N,\theta)\mathrm{logit}(\theta|mu,\sigma) / \mathrm{beta}(\theta|\mathrm{apos},\mathrm{bpos})$
on a log scale.  We take a simple brute force approach by evaluating $\log r(\theta)$ over the parameter space which is just $(0,1).$

```{r,echo=TRUE}
###set up a function for logr()

logr <- function(phi) {
  rtheta <- (sum(Y)*log(phi) + sum(N-Y)*log(1-phi) + 
               dlogitnorm(q=phi,mu=mu_prior,s=sigma_prior,log=TRUE))
              - dbeta(phi,shape1=apos,shape2=bpos,log=TRUE)
  return(rtheta)}

#evaluate logr over the (0,1) interval

theta_test <- seq(from=0.001,to=0.999,by=0.0001)
logrtheta<- logr(phi=theta_test)
summary(logrtheta)

plot(theta_test,logrtheta)

which.max(logrtheta)  ###This identifies the position in the logrtheta vector
                      ###at which the maximum value of logrtheta is                           ##### found

theta_maxr <- theta_test[which.max(logrtheta)]     #Value of theta which maximises logrtheta
logM2 <- logrtheta[which.max(logrtheta)]
theta_maxr
logM2
##check logM2 should equal the value of logr evaluated at theta_maxr
logr(theta_maxr)
```
 Now we know $\log M$ we can set up the rejection sampler.
```{r, echo=TRUE}
 ###Set up the rejection sampler
simnum <- 2000 #desired simulation sample size
storetheta2 <- vector(mode="numeric",length=simnum)
accepted2 <- 0
rejected2 <- 0

while(accepted2 < simnum) {
  theta_prop <- rbeta(n=1,shape=apos,shape2=bpos)
  ###Evaluate the log - importance ratio at this proposal
  logr_prop <- logr(phi=theta_prop)
  
  U <- runif(1)
  if (log(U) < logr_prop - logM2) {
    accepted2 <- accepted2 + 1
    storetheta2[accepted2] <- theta_prop
  } else {
    rejected2 <- rejected2 + 1
  }
}
summary(storetheta2)
plot(density(storetheta2))
accepted2
rejected2

acceptance_rate2 <- accepted2/(accepted2 + rejected2)
acceptance_rate2
##Superimpose density plot for beta approximation  densisty to 
#illustrate good approximator
plot(density(storetheta2),main="rejection sampler using beta density as the
       approximator",
     xlim=c(0,1),lwd=2)
lines(density_pos_betaapprx,lwd=2,col="red")
```

Next we repeat the beta-approximation approach for the model with 
$ \mathrm{logit-normal}(0,1)$ 
```{r,echo=TRUE}
mu_prior <- logit(0.5)
sigma_prior <- 1

##Re-draw the logit-normal  sample from the prior
testtheta <- rlogitnormal(n=10000,mu=mu_prior,sigma=sigma_prior)
density_prior <- density(testtheta)
plot(density(testtheta,from=0,to=1),main=paste("logitN(",round(mu_prior,2),",",sigma_prior,")",sep=""))
summary(testtheta)

###obtain a beta distribution with the same mean and variance as logit-normal sample
mean_theta <- mean(testtheta)
var_theta <- (sd(testtheta))^2
mean_theta
var_theta

a <- (mean_theta^2*(1-mean_theta) - var_theta*mean_theta)/var_theta 
b <- a*(1-mean_theta)/mean_theta
a
b

###plot the beta density with these values 
##could  do  exactly using dbeta but will simulate

apprxtheta <- rbeta(10000,a,b)
summary(apprxtheta)
mean(apprxtheta)
(sd(apprxtheta))^2
density_prior_beta <- density(apprxtheta,from=0,to=1)
plot(density_prior_beta,main=paste("beta(",round(a,2),",",round(b,2),")",sep=""))


###beta approximation to the posterior  -- 
#this will be the approximating density for the rejection sampler
## That is we are approximating the posterior with a distribution 
## that could plausibly  have been the posterior had a conjugate prior
##prior been adopted.

apos <- a + sumY #sum_i Y_i
bpos <- b + sumNmY  #sum_i (N_i-Y_i)

#check approximate posterior mean
apos/(apos+bpos)

pos_approx <- rbeta(10000,apos,bpos)
density_pos_betaapprx <- density(pos_approx)
plot(density(pos_approx))

####Need to find max of 
#Binomial(Y|N,\theta)logit(theta|mu,\sigma) / beta(\theta|apos,bpos);
# (on log scale)
##Function to compute logr
logr <- function(phi) {
  rtheta <- (sum(Y)*log(phi) + sum(N-Y)*log(1-phi) + 
               dlogitnorm(q=phi,mu=mu_prior,s=sigma_prior,log=TRUE))
  - dbeta(phi,shape1=apos,shape2=bpos,log=TRUE)
  return(rtheta)}

#evaluate logr over the (0,1) interval

theta_test <- seq(from=0.001,to=0.999,by=0.0001)
logrtheta<- logr(phi=theta_test)
summary(logrtheta)

plot(theta_test,logrtheta)

which.max(logrtheta)  ###This identifies the position in the logrtheta vector
###at which the maximum value of logrtheta is found

theta_maxr <- theta_test[which.max(logrtheta)]     #Value of theta which maximises logrtheta
logM2 <- logrtheta[which.max(logrtheta)]
theta_maxr
logM2
##check logM2 should equal the value of logr evaluated at theta_maxr
logr(theta_maxr)
```
Now, we know $\log M$ we can set up the rejection sampler.

```{r,echo=TRUE}

simnum <- 2000 #desired simulation sample size
storetheta3 <- vector(mode="numeric",length=simnum)
accepted3 <- 0
rejected3 <- 0

while(accepted3 < simnum) {
  theta_prop <- rbeta(n=1,shape=apos,shape2=bpos)
  ###Evaluate the log - importance ratio at this proposal
  logr_prop <- logr(phi=theta_prop)
  
  U <- runif(1)
  if (log(U) < logr_prop - logM2) {
    accepted3<- accepted3 + 1
    storetheta3[accepted3] <- theta_prop
  } else {
    rejected3 <- rejected3 + 1
  }
}
summary(storetheta3)
plot(density(storetheta3))
accepted3
rejected3

acceptance_rate3 <- accepted3/(accepted3 + rejected3)
acceptance_rate3
##Superimpose density plot for beta approximation  densisty to 
#illustrate good approximator
plot(density(storetheta3),main="rejection sampler using beta density as the
     approximator",
     xlim=c(0,1),lwd=2)
lines(density_pos_betaapprx,lwd=2,col="red")
```
