---
title: "Gibbs Sample for Zero-inflated Poisson model"
author: "Patrick Graham"
date: "4 October 2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Backround
We use an example from the Institute for Digital Research and Education at UCLA https://stats.idre.ucla.edu

The description given there is
``The state wildlife biologists want to model how many fish are being caught by fishers at a state park. Visitors are asked how long they stayed, how many people were in the group, were there children in the group and how many fish were caught. Some visitors do not fish, but there is no data on whether a person fished or not. Some visitors who did fish did not catch any fish so there are excess zeros in the data because of the people that did not fish.''

The data contain the response of 250 groups of park visitors who participated in the survey.

## Preliminaries
Set-up packages needed

```{r, echo=FALSE}
#library(ggplot2)
library(arm)
library(LearnBayes)
library(mvtnorm)
library(coda)  #use install.packages("coda") if not already installed
#Check  working directory and set if necessary

getwd()

#setwd("~/Patrick/314-2017/code) unhash and edit for a different working directory
```

Set up  a function for the Gelman-Rubin diagnostic and effective Monte Carlo sample size.

```{r, GRfunction}
## implement the chunked  version  of the Gelman-Rubin convergence diagnostic
## only  works  for  a scalar parameter and assumes the posterior sample
## is organised  as as npost by nchains matrix
##where  npost  is the simulation sample size after the burn-in has been
## discarded.

GRchunk <-  function(post){
  require(coda)
  
  possize <- nrow(post)
  
  n1 <- round(possize/2)  ##size of first chunk
  
  chunk1 <- post[1:n1,]
  chunk2 <- post[(round(possize/2)+1):possize,]
  
  post_chunked <- cbind(chunk1,chunk2)
  
    ##obtain the chain means
  chain_mean <- colMeans(post)
  ##obtain the within chain variance
  chain_sd <- apply(post,MARGIN=2,FUN=sd)
  chain_var <- chain_sd^2
  W = mean(chain_var)
  ##get between chain variance
  
  possize_chunked <- nrow(post_chunked)
  B <- possize_chunked*(sd(chain_mean))^2
  
  ##Now build the components of the Gelman-Rubin statistic
  
  Vplus <- ((possize_chunked-1)/possize_chunked) * W + (1/possize_chunked)*B
  Rhat <- sqrt(Vplus/W)
  Rhat
  
  ##compute  estimated effective sample size, using Coda
  
  post.mcmc <- coda::as.mcmc(post_chunked )
 
  codaNeff_chains  <- coda::effectiveSize(post.mcmc)
  
  codaNeff <- sum(codaNeff_chains)
  
  outlist <- list(Rhat=Rhat,codaNeff=codaNeff,codaNeff_chains=codaNeff_chains)
  return(outlist)

}
```

Read in the data and check the basic structure.
We assume the data is in a sub-folder of the working director called data

```{r,echo=TRUE}
fishdata <- read.csv("data/fish.csv") 
```

If the data are in the working directory unhash and run  the following commands

```{r, echo=TRUE}
#fishdata <- read.csv("data/fish.csv") #assuming the data  is in the current wd 
```
```{r, echo=TRUE}
str(fishdata)
```

set-up some variables as factors - this code fragment is direct from IDRE

```{r, echo=TRUE}
fishdata <- within(fishdata, {
  nofish <- factor(nofish)
  livebait <- factor(livebait)
  camper <- factor(camper)
})
```

Take a look at some basic summaries

```{r, echo=TRUE}
summary(fishdata)
```

For this example we will work with just the count variable. It is unclear  what the meaning  of the "nofish  variable is:"

```{r, echo=TRUE}


Y <- fishdata$count
N <- length(Y)   ###Number of observations

# histogram with x axis in log10 scale - add 0.1 to avoid log(0)

hist(log10(Y+0.1))
```

## A zero-inflated Poisson model for the counts

We let $Z$ denote the unobserved variable "fished" coded 1 if a visiting group fished and coded 0 if they did not.
We adopt the following model

$$
\begin{aligned}
Z|\phi \, & \sim \mathrm{Bernoulli}(\phi), \, \mathrm{independently} \,\,
\mbox{ for } i \,\, \mbox{ in } 1, \ldots, n \\
[Y|Z=1,\theta] \, & \sim \mathrm{Poisson}(\theta); \\
\Pr(Y=0|Z=0) & = 1 \\
\phi & \sim \mathrm{Beta}(a,b) \\
\theta &\sim \mathrm{Gamma}(c,d)
\end{aligned}
$$
Note that given a group does not fish, they will not catch any fish, hence $\Pr(Y=0|Z=0)=1.$

And we make the usual conditional independence assumptions. So, letting 
$\mathbf{Y}=(Y_{1},\ldots, Y_{N}),\, \mathbf{Z}=(Z_{1},\ldots, Z_{N})$

$$
\begin{aligned}
p(\mathbf{Z}|\phi) & =\prod_{i}\mathrm{Binomial}(Z_{i}|\phi) \\
p(\mathbf{Y}|\mathbf{Z},\theta) &= \prod_{i:Z_{i}=1}\mathrm{Poisson}(Y_{i}|\theta)
\end{aligned}
$$
For illustrative purposes we will use uninformative priors, setting $a=1,b=1$ and $c=0.01,d=0.01,$ giving
$$
\begin{aligned}
\phi & \sim \mathrm{Beta(1,1)} \\
\theta & \sim \mathrm{Gamma}(0.01,0.01)
\end{aligned}
$$

To help keep these prior parameters clear in the code below we rename the parameters of the prior for $\theta$ as follows

$$
\begin{aligned}
c &\rightarrow theta\_prior1 = 0.01 \\
d &  \rightarrow theta\_prior2 = 0.01
\end{aligned}
$$
and the parameters of the prior  for $\phi$ to
$$
\begin{aligned}
a &\rightarrow phi\_prior1 = 1 \\
b &  \rightarrow phi\_prior2 = 1
\end{aligned}
$$

##Gibbs-sampler

The unknowns in this problem are $(\mathbf{Z},\theta,\phi).$ We use the Gibbs sampler to compute
$p(\mathbf{Z},\theta,\phi|\mathbf{Y}).$ Primary interest is on $\theta$ -expected catch given that you fish; but $\phi$ - probability of fishing is also of some interest. Consequently our real aim is to compute
$$ p(\theta,\phi | \mathbf{Y})=\int p(\mathbf{Z},\theta,\phi|\mathbf{Y}) \, \mathrm{d}\mathbf{Z}. $$  Since $\mathbf{Z}$ is a vector of binary  varibles, the integral is really a sum over  all $\mathbf{Z}$ values.  

Having obtained obtained a sample from the joint posterior $p(\mathbf{Z},\theta,\phi|\mathbf{Y})$ via Gibbs sampling, the Monte Carlo equivalent to integrating over the latent $\mathbf{Z}$ values is simply to ignore them and use the sample of $\theta, \phi$ values for inference for $\theta$ and $\phi.$

The full conditional (conditional) posterior distributions for this problem are     
(i)
$$
\begin{aligned}
p(\theta|\mathbf{Z},\phi,\mathbf{Y}) & \propto p(\mathbf{Y}|\mathbf{Z}\theta)p(\theta) \\
& \propto \left [ \prod_{i:Z_{i}=1}\mathrm{Poisson}(Y_{i}|\theta) \right ] 
\mathrm{Gamma}(\theta|0.01,0.01) \\
& = \mathrm{Gamma}\left( \theta|\left (0.01+\sum_{i:Z_{i}=1}Y_{i} \right ),\left (0.01+\sum_{i:Z_{i}=1}1 \right )  \right )
\end{aligned}
$$
(ii)
$$
\begin{aligned}
p(\phi|\mathbf{Z},\theta,\mathbf{Y}) & \propto p(\mathbf{Y},\mathbf{Z} | \theta,\phi)p(\theta)p(\phi) \\
& = p(\mathbf{Y}|\mathbf{Z},\theta)p(\mathbf{Z}|\phi)p(\theta)p(\phi) \\
& \propto  p(\mathbf{Z}|\phi)p(\phi) \\
& = \left [\prod_{i}\mathrm{Bernoulli}(Z_{i}|\phi) \right ]\mathrm{Beta}(\theta|1,1) \\
& = \mathrm{Beta} \left (\theta |\left ( 1+\sum_{i}Z_{i} \right ), 
\left ( 1 + N-\sum_{i}Z_{i} \right ) \right )
\end{aligned}
$$
(iii) $p(\mathbf{Z}|\theta,\phi,\mathbf{Y})$     
Under our conditional independence modelling assumptions (given model parameters)
$$p(\mathbf{Z}|\theta,\phi,\mathbf{Y}) = \prod_{i}p(Z_{i}|\theta,\phi,Y_{i})).$$
That is, given $(\theta, \phi, \mathbf{Y})$ we can can generate the individual unobserved $Z$ values independently. There are really only two cases to  consider, depending on whether the observed count was greater than zero or equal to zero.

Recall $\Pr(Y=0|Z=0)=1$  since no fish can be caught if no fishing is done. 
Logically, if $Y > 1$  then it must be the case that $Z =1,$ (If fish were caught some fishing must have been done; that is 
\[
\Pr(Z=1|\theta,\phi,Y=y) = 1; \, y > 0 
\]
We can show this mathematically
$$
\begin{aligned}
\Pr(Z=1|Y=y,\theta,\phi) & = 1-\Pr(Z=0|Y=y,\theta,\phi) \\
                         & = 1- \frac{\Pr(Y=y|Z=0,\theta,\phi)\Pr(Z=0|\phi)}
                                      {\Pr(Y=y|Z=0,\theta,\phi)\Pr(Z=0|\phi)
                                      +\Pr(Y=y|Z=1,\theta)\Pr(Z=1|\phi)} \\
                         &  = 1 - 0; \,\, y > 0
\end{aligned}
$$
since $\Pr(Y=y|Z=0,\theta,\phi) = 0;\, y > 0.$

So if $Y_{i}> 0$ we should assign $Z_{i}=1.$ The only interesting case is therefore:
$$
\begin{aligned}
\Pr(Z=1|Y=0,\theta,\phi) & = \frac{\Pr(Y=0|Z=1|\theta)\Pr(Z=1|\phi)}
                                {\Pr(Y=0|Z=1|\theta)\Pr(Z=1|\phi) +
                                \Pr(Y=0|Z=0)\Pr(Z=0|\phi)} \\
                          & =  \frac{\mathrm{Poisson}(0|\theta)\phi}
                                {\mathrm{Poisson}(0|\theta)\phi +
                                1\times(1-\phi)}  \\
                          & = \frac{\exp(-\theta)\phi}
                                    {\exp(-\theta)\phi + (1-\phi)}
\end{aligned}
$$
Consequently to simulate $p(\mathbf{Z}|\theta,\phi,\mathbf{Y})$ we

1. set $Z_{i}=1$ if $Y_{i}=1$
2. draw $Z_{i}$ as $\mathrm{Bernoulli}\left (\frac{\exp(-\theta)\phi}
                                    {\exp(-\theta)\phi + (1-\phi)} \right )$
if $Y_{i}=0.$

Obviously the details of the derivations of the full conditional (conditional posterior distributions) are specific to this problem. However in problems of ``the missing data type" it does usually work out that the full conditionals for the parameters given the simulated missing data are straightforward to derive, whereas a little more work is required to  derive the full conditional for the missing data.

```{r,echo=FALSE}
##Optionally set seed for reproducibility
set.seed(29661)
```


```{r,echo=TRUE}
## set priors
theta_prior1 <- 0.01
theta_prior2 <- 0.01

phi_prior1 <- 1
phi_prior2 <- 1

nchains <- 5
nsim <- 2000
burnin <- 1000

store_phi <- matrix(NA,nrow=nsim,ncol=nchains)
store_theta <- matrix(NA, nrow=nsim,ncol=nchains)
```
Note we are only storing the generated values for $\phi$ and $\theta.$ We could also store $\mathbf{Z}$ but since the $Z$ values are not of intrinsic interest for this particular analysis  we do not store them.

```{r, echo=TRUE}
#starting values
##simple approach for theta - assume ordinary Poisson gamma model;
##unlikely to be that accurate

for (j in 1:nchains) {

  ##starting value for theta
theta <- rgamma(n=1,shape=sum(Y)+theta_prior1,rate=(N+theta_prior2))

##Could / should use a smaller N above to get an 
#overdispersed approximation

##No obvious approximation for beta so we just draw from the prior

##starting value for phi
phi <- rbeta(n=1,shape1=1,shape2=1)

 for (i in 1:nsim) {
   
  ##Simulate the Z values
   
   ###everyone with Y > 0 must have Z=1 so we only need to calculate
   ##obtain Pr(Z=1|Y=0)
   
   pZ_1 <- (dpois(0,theta)*phi) / (dpois(0,theta)*phi + 1-phi)
   
   pZ <- (Y>0) + (Y==0)*pZ_1  ##this is now a vector of length N
                              ###if (Y > 0) pZ=1; else pZ = pZ_1
   
   Z <- rbinom(n=N,prob=pZ,size=1)  ##vector of 0 or 1 indicating fishing 
                                    ##statu
   
   ##update theta  
   
   ##only learn about theta from the Z=1 group (fishers) so subset out this
   ##group
    Y1 <- Y[Z==1]
    sumY1 <- sum(Y1)
    
    theta <- rgamma(n=1,shape=(sumY1+theta_prior1),
                        rate=(length(Y1)+theta_prior2) )
   
   ###Update phi
   
   phi <- rbeta(n=1,shape1=(sum(Z)+phi_prior1),
                    shape2=(N-sum(Z)+phi_prior2) )
                 
  
  ###store phi and theta
  store_phi[i,j ] <- phi
  store_theta[i,j] <- theta
 }
}
```
Now check for  convergence

Discard burn-in samples:

```{r,echo=TRUE}
post_theta <- store_theta[(burnin+1):nsim,]
post_phi <-  store_phi[(burnin+1):nsim,]
```


### Gelman-Rubin diagnostics for theta

```{r, GRtheta}
GRtheta <- GRchunk(post_theta)

GRtheta$Rhat
GRtheta$codaNeff
GRtheta$codaNeff_chains
```

Gelman-Rubin diagnostics  for $\phi$

```{r, echo=TRUE}

GRphi <- GRchunk(post_phi)

GRphi$Rhat
GRphi$codaNeff
GRphi$codaNeff_chains



```
Convergence  and effective Monte Carlo  sample size look  very  good.  

Take a look at the trace plots

```{r, echo=TRUE}

#\theta
x <- seq(from=1,to=nsim,by=1)
plot(x,store_theta[,1],type="l")
lines(x,store_theta[,2],col="red")
lines(x,store_theta[,3],col="blue")
lines(x,store_theta[,4],col="green")
lines(x,store_theta[,5],col="orange")

# \phi
x <- seq(from=1,to=nsim,by=1)
plot(x,store_phi[,1],type="l")
lines(x,store_phi[,2],col="red")
lines(x,store_phi[,3],col="blue")
lines(x,store_phi[,4],col="green")
lines(x,store_phi[,5],col="orange")


###posterior density plots
plot(density(post_theta))

plot(density(post_phi))

##posterior summary statistics
probs <- c(0.025,0.5,0.975)

quantile(post_theta,probs)
##compare this with the naive MLE from a model that does not 
#take  account of the fact that some people do not fish (Z=0)
cat("naive MLE", (sum(Y)/N) )
```

Finally look at posterior inference for the proportion of visiting groups that fish

```{r, echo=TRUE}
quantile(post_phi,probs)
```

The data is reasonably informative about $\phi$ even though Z is not 
fully observed
